Adaptive Disk Storage for Interaction Graphs

### 2014-08-07 docs/cidr2015/text/main.tex
Moving paper into ACM SIG format.

\clearpage
\newpage
\section{Partitioning}
This section is still a work-in-progress. We are trying to find the algorithms.

@@

\clearpage
\newpage
\subsection{Overlapping Attributes}

@@

>>>>

\input{config}

%\documentclass{sig-alternate}

\begin{document}
%
% --- Author Metadata here ---
\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

\title{Adaptive Disk Storage for Interaction Graphs}

\numberofauthors{2} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
Robert Soul\'{e}\\
       \affaddr{University of Lugano}\\
       \affaddr{1932 Wallamaloo Lane}\\
       \affaddr{Wallamaloo, New Zealand}\\
       \email{robert.soule@usi.ch}
% 2nd. author
\alignauthor
Bu\u{g}ra Gedik\\
       \affaddr{Bilkent University}\\
       \affaddr{P.O. Box 1212}\\
       \affaddr{Dublin, Ohio 43017-6221}\\
       \email{bgedik@cs.bilkent.edu.tr}
}

\maketitle
\begin{abstract}
TODO
\end{abstract}

% A category with the (minimum) three required fields
\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

\terms{Theory}

\keywords{ACM proceedings, \LaTeX, text tagging}

\section{Introduction}

Query workload and disk layout have a significant impact on database
performance.  For table-based, relational databases, this fact has led database
designers to develop alternative approaches for storage layout: row-oriented
storage is more efficient when queries access many attributes from a small
number of records, and column-oriented storage is more efficient when queries
access a small number of attributes from many records.

Recently, there has been increased interest in graph databases in general, and
\emph{interaction graph} databases in particular.  An \emph{interaction graph}
is an append-only graph, where new edges and vertices are added as time
progresses.  Thus, interaction graph encode the relationships over time between
entities in their structure, making them ideal for use in social networking,
telecommunications, or modeling the world-wide web.

Unfortunately, although interaction graph databases, like relational databases,
are the target of diverse query workloads, there is no clear correspondence to a
row-oriented or column-oriented storage layout.
Typical interaction graphs have data (i.e., \emph{attributes} ) associated with
their vertices and edges. Most algorithms access this data as they traverse the
graph.

In an interaction graph database organized as blocks of temporal neighbor lists
(as in Gedik et al.~\cite{gedik14}), it is important to store the edge and
vertex attributes locally. For  instance, if the edge properties were kept separately
in a relational table, there will be almost no benefit to the locality
optimizations performed for block organization (discussed in Gedik et
al.~\cite{gedik14}, as we would have to go back and forth between the disk
blocks to access the edge attributes).%

On the other hand, putting all the edge attributes into the disk blocks
containing the graph structure is expected to cause significant overhead when
only a few of these attributes are read. This is somewhat similar to the

% However, not all of the data is
% accessed by all the queries. Typically, there are correlations among the
% attributes accessed by different queries, such as Q1 and Q5 accessing
% attributes a and b, and Q2, Q3, Q4 accessing attributes c and d, and so on.%

Typically, a graph database stores entities as vertices,
relationships as edges, and additional information in labels that may be
associated with either vertices or edges.

They store entities as vertices,
relationships as edges, and additional information in labels that may be
associated with either vertices or edges.

An \emph{interaction graph} is an append-only graph, where new edges and
vertices are added as time progresses.%

Typical interaction graphs have data associated with their vertices and edges.
We call this data \emph{attributes} or \emph{properties}. Most algorithms
access this data as they traverse the graph. However, not all of the data is
accessed by all the queries. Typically, there are correlations among the
attributes accessed by different queries, such as Q1 and Q5 accessing
attributes a and b, and Q2, Q3, Q4 accessing attributes c and d, and so on.%

% In an interaction graph database organized as blocks of temporal neighbor lists
% (as in Gedik et al.~\cite{gedik14}), it is important to store the edge and
% vertex properties locally. For  instance, if the edge properties are kept away
% in a relational table, there will be almost no benefit to the locality
% optimizations performed for block organization (discussed in Gedik et
% al.~\cite{gedik14}, as we would have to go back and forth between the disk
% blocks to access the edge attributes).%

On the other hand, putting all the edge attributes into the disk blocks
containing the graph structure is expected to cause significant overhead when
only a few of these attributes are read. This is somewhat similar to the 
problem with row-oriented databases, where the entire row needs to be accessed
from the disk despite the query needing only a small fraction of it.
Unfortunately, there is no clear correspondence to a column-oriented database
layout for the graph databases.%

An important characteristic of interaction graphs are that they are temporal.
As such, the co-access correlations for the attributes can be different for
different temporal regions. It might also be unknown at the insertion time and
may be discovered later depending on the workload. This points to the need for
adaptively optimizing the layout (somewhat similar to H2O \cite{alagiannis14}
and HYRISE~\cite{grund10}).%

I imagine a solution to this problem, which I name the `rail layout'. The idea
is to start with large blocks that contain the entire data. As we learn about
the access properties for different time regions, we might start splitting
such blocks into smaller blocks that run parallel to each other, almost like
having two or more graph databases for certain time regions, each containing a
 different subset of the attributes, but with a link between them in case a
query needs to access both.

\section{Problem Description}

The problem is to determine a partitioning of the disk blocks into sub-blocks
that would minimize query I/O. Of course, partitioning into sub-blocks results
in an increase in storage cost. The storage cost must stay below some
specified limit.

There are two possibilities: \emph{overlapping} and \emph{non-overlapping}
partitioning. In the overlapping case, attributes can be replicated in the
sub-blocks. The non-overlapping case is a true partition.

We start by defining the cost model, which formalizes the query I/O and storage
costs.

\subsection{Cost Model}
Let $Q$ be the query workload, where each query $q\in Q$ accesses a set of
attributes $q.A$ and traverses parts of the graph for the time range
$q.T=[q.t_s,q.t_e]$. We denote the set of all attributes as $A$. Given a block
$B$, we denote its time range as $B.T$, which is the union of the time ranges
of its temporal neighborlists. Let $s(a)$ denote the size of an attribute $a$.
We use $c_n(B)$ to denote the number of temporal neighborlists within block
$B$ and $c_e(B)$ to denote the total number of edges in the temporal
neighborlists within the block. We overload the notation for block size and
use $s(B)$ to denote the size of a block $B$. We have: 
\begin{equation}
s(B) = c_e(B) \cdot \Big(16 + \sum_{a\in B.A} s(a)\Big) + c_n(B) \cdot 12  
\end{equation}
Here, $16$ corresponds to the cost of storing the edge id and the timestamp,
and $12$ corresponds to the cost of storing the head vertex ($8$ bytes) plus
the number of entries ($4$ bytes) for a temporal neighborlist.

Our goal is to create a potentially overlapping partitioning of a block,
denoted by $\mathcal{P}(B)$. We call the partitions \emph{sub-blocks}. We have
$\bigcup_{B'\in \mathcal{P}(B)} B'.A = A$. We aim to find the function
$\mathcal{P}$ that minimizes the query I/O over $B$, while keeping the
relative storage overhead beyond a limit, say $1+\alpha$ times the original.
If we represent the query I/O as $L(\mathcal{P}, B)$ and the relative storage
overhead as $H(\mathcal{P}, B)$, our goal is to find:
\begin{equation}
\mathcal{P} \leftarrow \mbox{argmin}_{\{\mathcal{P}: H(\mathcal{P}(B)) < \alpha\}} L(\mathcal{P},B)
\end{equation}

The storage overhead can be formalized as follows for the non-overlapping case:
\begin{equation}
H(\mathcal{P}, B) = (|\mathcal{P}(B)|-1)\cdot\Big(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\Big) 
\end{equation}

For the general case (including overlapping), we can formulate the storage
overhead as follows:
\begin{equation}
H(\mathcal{P}, B) = \frac{\sum_{B'\in \mathcal{P}(B)} s(B')}{s(B)} - 1 
\end{equation}

%TODO: note about why we have two different cases

\subsubsection{Query I/O}

Let $m$ be a function that maps a query $q$ to the set of sub-blocks that are
accessed to satisfy it for a relevant block $B$ under a given partitioning
$\mathcal{P}$. For the case of non-overlapping attributes, we have:
\begin{equation}
m(\mathcal{P}, B, q) = \{B': B'\in \mathcal{P}(B) \wedge q.A \cap B'.A \ne \emptyset\}  
\end{equation}

For the case of overlapping attributes, we use a simple heuristic to define the
set of sub-blocks to be used for answering the query.
Algorithm~\ref{alg:greedyM} captures it. The idea is to select sub-blocks that
bring the highest relative marginal gain, which is defined as the size of
attribute data that contributes to the query result, relative to the sub-block
size. While computing the marginal gain, attributes that are covered by
sub-blocks that are already selected are not considered.
%
\begin{algorithm}[ht]
\scriptsize
\caption{m-overlapping($\mathcal{P}, B, q$)}
\label{alg:greedyM}
\KwData{$\mathcal{P}$: partitioning function, $B$: block, $q$: query}
$S\leftarrow \emptyset; R\leftarrow \emptyset$ \tcp*{Selected attributes; Resulting sub-blocks}
\While(\tcp*[f]{While unselected attributes remain}){$S \subset q.A$}{
  $B' \leftarrow \mbox{argmax}_{B'\in\mathcal{P}(B)\setminus R} \sum_{a\in B'.A \cap q.A \setminus S} \frac{c_e(B') \cdot s(a)}{s(B')}$
  $S \leftarrow S \cup B'.A$\tcp*{Extend the selected attributes}
  $R\leftarrow R \cup B'$\tcp*{Extend the selected sub-blocks}
}
\Return R \tcp*{Final set of sub-blocks covering the query attributes}
\end{algorithm} 
%
The query I/O cost for a block is then given by:
\begin{equation}
L(\mathcal{P}, B) = \sum_{q\in Q} \mathbf{1}(q.T \cap B.T \neq \emptyset) \cdot \sum_{B'\in m(\mathcal{P}, B, q)} \!\!s(B')
\end{equation}

%\paragraph{\textbf{Other useful equations}}
%\noindent The following are unused, but may be needed for the partitioning
%algorithm.
%
%Let $f: \mathbb{R}\times A \rightarrow \mathbb{N}$ be a function that maps a
%attribute-timestamp pair $(a, t)$ to the frequency of queries that access
%attribute $a$ for time $t$. More formally:
%\begin{equation}
%f(a,t) = \sum_{q\in Q} \textbf{1}(a\in q.A \wedge t\in q.T)
%\end{equation}
%
%Let $T$ represent a time range $[T.s, T.e]$. We denote the frequency of
%attribute access for the time interval $T$ for attribute $a$ as follows:
%\begin{equation}
%f(a, T) = \int_{T.s}^{T.e} f(a, t)\cdot dt =  \sum_{q\in Q} |q.T \cap T| 
%\end{equation}

\section{Partitioning}
This section is still a work-in-progress. We are trying to find the algorithms.

@@

\subsection{Overlapping Attributes}

@@

\section{Related Work}
\section{Conclusion}

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{main}  % sigproc.bib is the name of the Bibliography in this case

\end{document}
